Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2016-02-02 WSSSPE3/report/WSSSPE3_review.tex
Section 4: Beautified.

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}} The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it isn't well understood yet.
Birgit Penzenstadler argued that we often define sustainability too narrowly.
Instead, sustainability at its heart is a systemic concept and must be viewed
from a range of different dimensions including environmental, economic,
individual, social, and technical. She introduced the Karlskrona Manifesto on
Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
of related work on the topic of sustainability into a set of (mis-)perceptions
and principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather commitments to rethink, to move
beyond the silo mentality, and to analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

@@

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with CIG: the Computational
Infrastructure for Geodynamics, a community software with a worldwide user base.
Like others, their main goals include high usability, sustainability, and
reproducibility. In order to achieve these goals, communication channels have
been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
and webinars. In order to contribute to the infrastructure, codes must adhere to
specified minimum standards with the desire that all codes are working toward
target standards.
%, but ideals are also given to encourage reaching the aforementioned goals. 
These include, e.g., the use of version control, certain coding styles, the
presence and nature of code tests and documentation, and certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
Hwang mentioned various ways to cite efforts within the framework, including
science papers, code papers, user manuals, and the CIG website. An analysis of
the resulting citations showed that $80\%$ of papers that use CIG codes mention
the code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used. Compared to other codes, CIG seems to be much better cited. In
part, this is attributed to the fact that CIG requires that donated software
provide a citable paper specified in the User Manual. The project is working on
tools and methods to generate attribution information automatically.
%This,
%in part, is attributed to suggested citation forms communicated to the
%community, and runtime tools that help collecting a relevant citation list
%automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the high energy physics effort to
biology, astrophysics, digital curation, and other disciplines. It includes
physicists, digital librarians, as well as computer scientists, and aims to
achieve some commonality across disciplines. Examples are meta-data descriptions
of archived data, computational descriptions, descriptions of how data was
processed, questions such as whether computation replication can be automated,
and what the impact of access policies on the preservation infrastructure is.
One of the products of this effort is a suite of tools that deals with this
preservation, and the questions was posed whether that software itself is
sustainable. Points that were brought up included the need for a user community
depending on a given software, and the need to provide added value for its
users. An important method to achieve this was to work with the user community
from the start, and to budget that way. For the specific example of the
preservation software, this means that besides adding value to the community, it
needs to be transparent to their workflows, i.e., not requiring additional
effort to preserve ``research objects''.

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}} James
Hetherington started by listing some of the problems research software faces,
which include low levels of reuse and poor standard of verification. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level. Hetherington
hypothesized that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for researchers who don't want to be programmers, and synergies
with research computing platforms. The success of such a group would be measured
by the output and quality of the research software, and could be part of
Research Computing or faculty. Members would not be independent researchers, but
would have to have a research background. An attempt to form a community of RSE
groups within the UK has been underway for several years, including funding from
the UK Research Council. An open question is whether this approach can be
adopted in other countries, including the USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, great developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design, development, and testing phase. The main
deficiency with this approach was pointed out to be the too-late interaction
with users. Instead, an alternative approach was proposed that, at first, skips
the development phase and repeatedly goes through requirement gathering, design
and user interaction/learning phases, and only eventually starts development
once an agreement is reached, leading to an increased user satisfaction, higher
adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
Platform for Predictive Biology: Lessons Learned on the KBase Project}} In this
talk, Dan Gunter presented experiences and lessons learned as part of KBase, an
open software and data platform for addressing the grand challenge of systems
biology: predicting and designing biological function. KBase is a unified system
that integrates data and analytical tools for comparative functional genomics of
microbes, plants, and their communities. However, it is also a collaborative
environment for sharing methods and results, and placing those results in the
context of knowledge in the field. Being a large, multi-institutional project,
one of the big challenges is to agree on standards to enable a single,
maintainable system. Working in isolation does not work (anymore) within this
field, and the community in the field also does not have standards for software
engineering. This is contrasted to computer science research, where software
engineering standards shorten design cycles, leading to more time for highly
rewarded activities like publishing, performance studies, graduating students,
or protecting ideas before publication. Instead of this more traditional
approach, KBase uses a variation on the ``Scrum'' methodology. After picking
projects and team members, four to five teams work on projects for about two
weeks before a one-week evaluation by the executive committee. Based on their
assessment new teams and projects might be chosen before iteratively restarting
the ``agile'' development cycle. This process is intentionally open and
documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: disconnects between
experimental data, research software, and publications. A current effort in the
Geosciences is the ``Geoscience Paper of the Future,'' which includes four
elements. First, it forms a modern paper including text, data, and pointers to
supplementary materials. Second, it is reproducible, including data processing,
workflow, and visualization tools. Third, it is part of open science, which
includes being publicly available under open licenses, and providing meta-data.
Last, but not least, it uses digital scholarship elements like persistent
identifiers for software, data, and authors, and it cites both data and
software. OntoSoft is a tool to help this effort, providing software stewardship
for the Geosciences. As part of this initiative, a special issue of a journal in
Geoscience areas is planned to include only Geoscience papers of the future,
with submissions open until the end of 2015. In addition, training sessions are
provided to geoscientists to learn best practices in software and data sharing,
provenance documentation, and scholarly publication.

@@

\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
This talk highlighted an attempt to solve the common problem for open source
development: it is difficult to collect information about how many people use
particular software, how often, which version, which parts of the software, or
on which operating system. Current solutions include surveys, but these have
high uncertainties. A different approach is based on automatic ``call-backs''
that collect these information at runtime and send it to a central place for
analysis. Problems with this approach include obtaining agreement from the user,
legal issues, increased maintenance (servers), ethical questions, and also the
lack of a common infrastructure. Some of these problems are of a social nature
and have to be solved as such, but the last problem (a missing common
infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
which is developing common APIs and a library implementation for common,
repeating tasks such as obtaining user consent. Results are uploaded not
directly to project servers, but to neutral third parties that only publish
aggregated statistics to projects.

@@

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
Developing Community}} The effort behind this talk by Ray Idaszak started from a
break-out group at an NSF SI2 workshop in 2015, and centers around creating a
framework for creation of metrics for scientific software. This framework would
improve both the metrics and the software it evaluates, and could also serve as
a tool for building a community around the idea. With especially the latter idea
(building a community) in mind, a software ``peer review group'' would be
created, representing stakeholders who will self-review software created by
their respective communities, and will concurrently develop metrics. The whole
project should be community-governed, without a single institution overseeing
the activities or infrastructure, with the hope to evolving community-generated
and adopted standards. The generation of metrics would be tied to the actual
evaluation of software, creating an incentive by improving the evaluated
software itself during this process. The framework code would provide
infrastructure for the creation of metrics and evaluation, and forums for
generation of software success metrics. It would also support code reviews of
the evaluated software. An open question is whether it is possible to fit the
resulting metrics in a common template. So far, this is still in a design phase,
with a white paper at the CSESSP workshop 2015 and this talk, but the WSSSPE
workshops are seen as a forum for the community to assemble and act, and is
planned to be used also in the future to build this community and framework.

>>>>

\item \textbf{Birgit Penzenstadler, Colin Venters, Christoph Becker, Stefanie
Betz, Ruzanna Chitchyan, Let\'{i}cia Duboc, Steve Easterbrook, Guillermo
Rodriguez-Navas, and Norbert Seyff: \textit{Manifesting the Ghost of the Future:
Sustainability}} The concept of sustainability has become a topic of interest in
the field of computing, which is evidenced by the increase in the number of
events that focus on the topic. Nevertheless, it is not well understood yet.
Birgit Penzenstadler argued that we often define sustainability too narrowly.
Instead, sustainability at its heart is a systemic concept and must be viewed
from a range of different dimensions including environmental, economic,
individual, social, and technical. She introduced the Karlskrona Manifesto on
Software Design~\cite{Becker:2014}, which distills knowledge from a broad range
of related work on the topic of sustainability into a set of (mis-)perceptions
and principles. The manifesto does not proclaim that there is an easy,
one-size-fits-all solution around the corner, but rather points out that
sustainability is a ``wicked problem'' and is often misunderstood. Due to these
misperceptions, even though sustainability's importance is increasingly
recognized, many software systems are unsustainable. Even more alarming is that
most software systems' broader impacts on sustainability are unknown. To change
this, the Karlskrona Manifesto proposes nine principles and commitments. These
commitments are not dogmatic laws, but rather commitments to rethink, to move
beyond the silo mentality, and to analyze in more depth. As such, they do not
restrict, but rather open up a space for discussion.

@@

\item \textbf{Louise Kellogg and Lorraine Hwang: \textit{Advancing Earth Science
through Best Practices in Open Source Software: Computational Infrastructure for
Geodynamics}} Lorraine Hwang presented experiences with the Computational
Infrastructure for Geodynamics (CIG), a community software with a worldwide user base.
Like others, their main goals include high usability, sustainability, and
reproducibility. As a means to achieve these goals, various communication channels have
been developed, such as mailing lists, wikis, workshops, hackathons, tutorials,
and webinars. In order to contribute to the infrastructure, codes must adhere to
specified minimum standards with the desire that all codes are working toward
target standards.
%, but ideals are also given to encourage reaching the aforementioned goals. 
These include, e.g., the use of version control, certain coding styles, the
presence and nature of code tests and documentation, and certain user workflows.

\item \textbf{Lorraine Hwang, Joe Dumit, Alison Fish, Louise Kellogg, Mackenzie
Smith, and Laura Soito: \textit{Software Attribution for Geoscience Applications
in the Computational Infrastructure for Geodynamics}} In a second talk, Lorraine
Hwang mentioned a variety of ways to cite efforts within the framework, including
science papers, code papers, user manuals, and the CIG website. An analysis of
the resulting citations showed that $80\%$ of papers that use CIG codes mention
the code name, and about the same number includes a citation. Only about $20\%$
acknowledge CIG. Within the same sample of papers, about one fifth use an URL to
cite codes (including non-CIG codes), and only about one eighth specify the
version used. Compared to other codes, CIG seems to be much better cited. In
part, this is attributed to the fact that CIG requires that donated software
provide a citable paper specified in the User Manual. The project is working on
tools and methods to generate attribution information automatically.
%This,
%in part, is attributed to suggested citation forms communicated to the
%community, and runtime tools that help collecting a relevant citation list
%automatically.

\item \textbf{Mike Hildreth, Jarek Nabrzyski, Da Huo, Peter Ivie, Haiyan Meng,
Douglas Thain, and Charles Vardeman: \textit{Data And Software Preservation for
Open Science (DASPOS)}} DASPOS is an NSF-funded multi-disciplinary effort,
located at Notre Dame and Chicago, that links the high energy physics effort to
other disciplines such as biology, astrophysics, and digital curation. It includes
physicists, digital librarians, as well as computer scientists, and aims to
achieve some commonality across disciplines. Examples are meta-data descriptions
of archived data, computational descriptions, descriptions of how data was
processed, questions such as whether computation replication can be automated,
and what the impact of access policies on the preservation infrastructure is.
One of the products of this effort is a suite of tools that deals with this
preservation, and the questions was posed whether that software itself is
sustainable. Points that were brought up included the need for a user community
depending on a given software, and the need to provide added value for its
users. An important method to achieve this was to work with the user community
from the start, and to budget that way. For the specific example of the
preservation software, this means that besides adding value to the community, it
needs to be transparent to their workflows, i.e., not requiring additional
effort to preserve ``research objects.''

\item \textbf{James Hetherington, Jonathan Cooper, Robert Haines, Simon
Hettrick, James Spencer, Mark Stillwell, Mike Croucher, Christopher Woods, and
Susheel Varma: \textit{An update from UK Research Software Engineers}} James
Hetherington started by listing some of the problems research software faces,
which include poor standard of verification and  low levels of reuse. For a long
time, technical solutions to such technical problems were focus of the eResearch
community, including research software distributions, grids, middleware and
workflows. Some limited adoption can be seen today in research communities, but
the main problems have not been solved to a sufficient level. Hetherington
hypothesized that instead of technical solutions, social innovation is needed: a
new role in the academic system focused on research software that combines the
best parts of a craftsperson and a scholar. However, social innovation in
centuries-old institutions is hard. Alternatives to such a new role would have
to include rewards for good research software, recognition of software as
academic output, and rejecting submissions based on irreproducible computational
results. Some advantages of research software engineering (RSE) groups include
the possibility of training in reproducible computational research, providing
collaborations for researchers who do not want to serve as programmers, and creating synergies
with research computing platforms. The success of such a group would be measured
by the output and quality of the research software. Members could be part of
Research Computing or faculty. They would not be independent researchers, but
would have to have a research background. An attempt to form a community of RSE
groups within the UK has been underway for several years, including funding from
the UK Research Council. An open question is whether this approach can be
adopted in other countries, including the~USA.

\item \textbf{Dan Gunter, Sarah Poon, and Lavanya Ramakrishnan: \textit{Bringing
the User into Building Sustainable Software for Science}} Dan Gunter's main
question was, ``What is needed to develop sustainable software?''. Beyond the
usual suspects of funding, proficient developers, good design, and software
engineering practices, Dan placed the users. He explained this using a
traditional software-development model starting from gathering requirements, and
reaching release through a design, development, and testing phase. The main
deficiency with this approach was pointed out to be the too-late interaction
with users. Instead, an alternative approach was proposed that, at first, skips
the development phase and repeatedly goes through requirement gathering, design
and user interaction/learning phases, and only eventually starts development
once an agreement is reached, leading to an increased user satisfaction, higher
adoption, and eventually to sustained software.

\item \textbf{Dan Gunter, Adam Arkin, Rick Stevens, Robert Cottingham, Sergei
Maslov, and the KBase Project: \textit{Challenges of a Sustainable Software
Platform for Predictive Biology: Lessons Learned on the KBase Project}} 
Dan Gunter presented experiences and lessons learned as part of KBase, an
open software and data platform for addressing the grand challenge of systems
biology: predicting and designing biological function. KBase is a unified system
that integrates data and analytical tools for comparative functional genomics of
microbes, plants, and their communities. However, it is also a collaborative
environment for sharing methods and results, and placing those results in the
context of knowledge in the field. Being a large, multi-institutional project,
one of the big challenges is to agree on standards to enable a single,
maintainable system. Working in isolation does not work (anymore) within this
field, and the community in the field also does not have standards for software
engineering. This is contrasted to computer science research, where software
engineering standards shorten design cycles, leading to more time for highly
rewarded activities like publishing, performance studies, graduating students,
or protecting ideas before publication. Instead of this more traditional
approach, KBase uses a variation on the ``Scrum'' methodology. After picking
projects and team members, four to five teams work on projects for about two
weeks before a one-week evaluation by the executive committee. Based on their
assessment new teams and projects might be chosen before iteratively restarting
the ``agile'' development cycle. This process is intentionally open and
documented.

\item \textbf{Yolanda Gil, Chris Duffy, Chris Mattmann, Erin Robinson, and Karan
Venayagamoorthy: \textit{The Geoscience Paper of the Future Initiative: Training
Scientists in Best Practices of Software Sharing}} Erin Robinson presented an
approach to overcome some hurdles in scientific publishing: disconnects among
experimental data, research software, and publications. A current effort in the
Geosciences is the ``Geoscience Paper of the Future,'' which includes four
elements. First, it forms a modern paper including text, data, and pointers to
supplementary materials. Second, it is reproducible, including data processing,
workflow, and visualization tools. Third, it is part of open science, which
includes being publicly available under open licenses, and providing meta-data.
Last but not least, it uses digital scholarship elements like persistent
identifiers for software, data, and authors, and it cites both data and
software. OntoSoft is a tool for helping with this effort, providing software stewardship
for the Geosciences. As part of this initiative, a special issue of a journal in
Geoscience areas is planned to include only Geoscience papers of the future,
with submissions open until the end of 2015. In addition, training sessions are
provided to geoscientists to learn best practices in software and data sharing,
provenance documentation, and scholarly publication.

@@

\item \textbf{Matthias Bussonnier: \textit{User Data Collection in Open Source}}
This talk highlighted an attempt to solve the common problem for open source
development: it is difficult to collect information about how many people use
particular software, how often, which version, which parts of the software, or
on which operating system. Current solutions include surveys, but these have
high uncertainties. A different approach is based on automatic ``call-backs''
that collect these information at runtime and send it to a central place for
analysis. Problems with this approach include obtaining agreement from the user,
legal issues, increased maintenance (of servers), ethical questions, and also the
lack of a common infrastructure. Some of these problems are of a social nature
and have to be solved as such, but the last problem (a missing common
infrastructure) is attacked by the sempervirens project~\cite{sempervirens},
which is developing common APIs and a library implementation for common,
repeating tasks such as obtaining user consent. Results are uploaded not
directly to project servers, but to neutral third parties that only publish
aggregated statistics to projects.

@@

\item \textbf{Stan Ahalt, Bruce Berriman, Maxine Brown, Jeffrey Carver, Neil
Chue Hong, Allison Fish, Ray Idaszak, Greg Newman, Dhabaleswar Panda, Abani
Patra, Elbridge Gerry Puckett, Chris Roland, Douglas Thain, Selcuk Uluagac, and
Bo Zhang: \textit{Scientific Software Success: Developing Metrics While
Developing Community}} The effort behind this talk given by Ray Idaszak started from a
breakout group at an NSF SI2 workshop in 2015, and centers around building a
framework for creating metrics for scientific software. This framework would
improve both the metrics and the software it evaluates, and could also serve as
a tool for building a community around the idea. With especially the last-mentioned idea
(building a community) in mind, a software ``peer review group'' would be
created, representing stakeholders who will self-review software created by
their respective communities, and will concurrently develop metrics. The whole
project should be community-governed, without a single institution overseeing
the activities or infrastructure, with the hope to evolving community-generated
and adopted standards. The generation of metrics would be tied to the actual
evaluation of software, creating an incentive by improving the evaluated
software itself during this process. The framework code would provide
infrastructure for the creation of metrics and evaluation, and forums for
generation of software success metrics. It would also support code reviews of
the evaluated software. An open question is whether it is possible to fit the
resulting metrics in a common template. So far, this is still in a design phase,
with a white paper at the CSESSP workshop 2015 and this talk, but the WSSSPE
workshops are seen as a forum for the community to assemble and act, and is
planned to be used also in the future to build this community and framework.
